### 2023-11-19 sections/2-methods.tex
start of rho-NN overlapping children

For a cluster $C$ with $|C|$ points, we begin by taking a random sample of $\sqrt{|C|}$ of its points, and computing pairwise distances between all points in the sample.
Based on those pairwise distances, we compute the \emph{geometric median} of this sample; 
i.e., the point which minimizes the sum of distances to all other points in the sample.
This geometric median is $C$'s \emph{center}.
The \emph{radius} of $C$ is the maximum distance from any point in $C$ to its center.
The point $l$ which is responsible for that radius (i.e., the furthest point from the center) is designated the \emph{left pole}.
The point $r$ which is furthest from $l$ is designated the \emph{right pole}.
We then partition the cluster into a \emph{left child} and \emph{right child}, where the left child contains all points in the cluster which are closer to $l$ than to $r$, and the right child contains all points in the cluster which are closer to $r$ than to $l$.
Without loss of generality, we assign to the left child those points which are equidistant from $l$ and $r$.
Starting from a root-cluster containing the entire dataset, we repeat this procedure until each leaf contains only one datum or some other user-specified stopping criterion is met.
This process is described in Algorithm \ref{alg:methods:clustering:partition}.

@@

\begin{itemize}
    \item $\delta = f(q, c)$ is the distance from the query to the cluster center $c$.
    \item $\delta_{max} = \delta + r$ is the distance from the query to the theoretically farthest possible instance in $C$.
    \item $\delta_{min} = \text{max}(0, \delta - r)$ is the distance from the query to the theoretically closest possible instance in $C$.
\end{itemize}

We define a \emph{singleton} as a cluster which either contains a single point (i.e., has cardinality 1) or which contains 
duplicates of the same point (i.e., has cardinality greater than 1 but contains only one unique point).
A singleton clearly has zero radius, and so $\delta_{max} = \delta = \delta_{min}$ for these clusters.
Hence, we sometimes overload the above notation to also refer to distance from a query to a point;
in these cases, we also have that $\delta_{max} = \delta = \delta_{min}$.

@@

We conduct $\rho$-nearest neighbors search as described in \cite{ishaq2019clustered}, but with the following improvement:
when a cluster overlaps with the query ball, instead of  always proceeding with both of its children, we proceed only with those children which also overlap with the query ball.
This improvement is described in Algorithm \ref{alg:methods:rnn-search}.

\begin{algorithm} 
    \caption{$\rho$-NN(\emph{cluster, query, r})} 
    \label{alg:methods:rnn-search} 
    \begin{algorithmic}
        \REQUIRE $r \geq 0$
        \REQUIRE $cluster \neq \emptyset$
        \STATE $results \Leftarrow \emptyset$
        \IF{$cluster.left \cap B_X(query, r) \neq \emptyset$}
            \IF{$cluster.left.\delta$ $\leq$ $r$ + $cluster.left.radius$}
                \STATE $\rho$-NN($cluster.left, query, r$)
            \ENDIF
        \ENDIF
        \IF{$cluster.right \cap B_X(query, r) \neq \emptyset$}
            \IF{$cluster.right.\delta$ $\leq$ $r$ + $cluster.right.radius$} 
                \STATE $\rho$-NN($cluster.right, query, r$)
            \ENDIF
        \ENDIF
        \IF{$cluster.isLeaf$}
            \FOR{$p \in cluster.points$}
                \IF{$p.\delta \leq r$}
                    \STATE $results.push((p, r))$
                \ENDIF
            \ENDFOR
        \ENDIF
        \STATE Return $results$
    \end{algorithmic}
\end{algorithm}

@@

First, we aim to find the element $q_{\tau} \in Q$ with the smallest $\delta_{max}$ such that 
$q_{\tau}$ and all elements in $Q$ with smaller $\delta_{max}$ collectively contain at least $k$ points.

To find this element, we use a process similar to the Partition algorithm used in Quicksort~\cite{10.1093/comjnl/5.1.10}, adjusted to account for the varying cardinalities of clusters in $Q$ by having each cluster count a number of times equal to its cardinality.
Our modified version of this algorithm finds the smallest index $i$ such that all elements in $Q$ with $\delta_{max}$ less than or equal to $Q[i]$'s $\delta_{max}$ collectively have cardinality greater than or equal to $k$.
We consider $Q[i]$ to be our $q_{\tau}$, and our threshold $\tau$ to be $Q[i]$'s $\delta_{max}$.

We then remove from $Q$ any element whose $\delta_{min}$ is greater than $\tau$.
Then we remove all leaf clusters from $Q$ and add their points to $Q$ instead.
Finally, we replace all non-leaf clusters in $Q$ with their children.

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Sieve(\emph{root, query, k})} % give the algorithm a caption
    \label{alg:methods:sieve} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \REQUIRE $root$, $query$, $k$
        \REQUIRE $root.cardinality \geq k$
        \STATE $Q \Leftarrow$ [$root$]
        \WHILE{$|Q| > k$}
            \STATE $i \Leftarrow QuicksortPartition(Q, k)$
            \STATE $\tau \Leftarrow Q[i].\delta_{max}$
            \FOR {$q \in Q$}
                \IF {$q.\delta_{min} > \tau$}
                    \STATE $Q.pop(q)$
                \ENDIF
            \ENDFOR
            \FOR {$q \in Q$}
                \IF {$q.isLeaf \lor |q| \leq k$}
                    \STATE $Q.append(q.points)$
                \ELSE
                    \STATE $[l, r] \Leftarrow q.children$
                    \STATE $Q.append([l, r])$   
                \ENDIF
                \STATE $Q.remove(q)$
            \ENDFOR 
        \ENDWHILE
        \STATE Return $Q$
    \end{algorithmic}
\end{algorithm}

@@

We first let $Q$ be a min-queue of clusters prioritized by $\delta_{min}$.
$Q$ starts containing only the root cluster.
We also let $H$ be a fixed-size ($k$) max-queue of points prioritized by $\delta$.
$H$ starts empty.

While $H$ has fewer than $k$ points or $\delta$ of the top priority element in $H$ is greater than $\delta_{min}$ of the top priority element in $Q$, we do the following:

@@

\begin{algorithm} 
\caption{Greedy Sieve(\emph{root, query, k})} 
\label{alg:greedy-sieve} 
\begin{algorithmic}
    \STATE $Q \Leftarrow$ priority queue
    \STATE $Q.push(root)$
    \STATE $H \Leftarrow$ priority queue
    \WHILE{$|H| < k \lor H.peek.\delta > Q.peek.\delta_{min}$}
        \WHILE{$\neg Q.peek.isLeaf$}
            \STATE $[l, r] \Leftarrow Q.pop().children$
            \STATE $Q.push([l, r])$
        \ENDWHILE
        \STATE $leaf \Leftarrow Q.pop()$
        \STATE $H.push(leaf)$
        \WHILE{$|H| > k$}
            \STATE $H.pop()$
        \ENDWHILE
    \ENDWHILE
    \STATE Return $H$
\end{algorithmic}
\end{algorithm}

>>>>

For a cluster $C$ with $|C|$ points, we begin by taking a random sample of $\sqrt{|C|}$ of its points, and computing pairwise distances between all points in the sample.
Based on those pairwise distances, we compute the \emph{geometric median} of this sample; 
i.e., the point which minimizes the sum of distances to all other points in the sample.
This geometric median is $C$'s \emph{center}.
The \emph{radius} of $C$ is the maximum distance from any point in $C$ to its center.
The point $\ell$ which is responsible for that radius (i.e., the furthest point from the center) is designated the \emph{left pole}.
The point $r$ which is furthest from $\ell$ is designated the \emph{right pole}.
We then partition the cluster into a \emph{left child} and \emph{right child}, where the left child contains all points in the cluster which are closer to $\ell$ than to $r$, and the right child contains all points in the cluster which are closer to $r$ than to $\ell$.
Without loss of generality, we assign to the left child those points which are equidistant from $\ell$ and $r$.
Starting from a root-cluster containing the entire dataset, we repeat this procedure until each leaf contains only one datum or some other user-specified stopping criterion is met.
This process is described in Algorithm \ref{alg:methods:clustering:partition}.

@@

\begin{itemize}
    \item $\delta = f(q, c)$ is the distance from the query to the cluster center $c$.
    \item $\delta_{\max} = \delta + r$ is the distance from the query to the theoretically farthest possible instance in $C$.
    \item $\delta_{\min} = \text{max}(0, \delta - r)$ is the distance from the query to the theoretically closest possible instance in $C$.
\end{itemize}

We define a \emph{singleton} as a cluster which either contains a single point (i.e., has cardinality 1) or which contains 
duplicates of the same point (i.e., has cardinality greater than 1 but contains only one unique point).
A singleton clearly has zero radius, and so $\delta_{\max} = \delta = \delta_{\min}$ for these clusters.
Hence, we sometimes overload the above notation to also refer to distance from a query to a point;
in these cases, we also have that $\delta_{\max} = \delta = \delta_{\min}$.

@@

We conduct $\rho$-nearest neighbors search as described in~\cite{ishaq2019clustered}, but with the following improvement:
when a cluster overlaps the query ball, instead of always proceeding with both of its children, we proceed only with those children which also overlap the query ball.

To determine whether both children overlap the query ball, we consider the following geometry. 
Let $c$ denote the center of a cluster, $\ell$ and $r$ its left and right poles (see Section~\ref{subsubsec:methods:cluster-partitioning}), $q$ the query, and $\rho$ the search radius, as shown in Figure [ADD FIGURE].
Without loss of generality, we assume that $\overline{qr} \leq \overline{q\ell}$ (i.e., the query is closer to the right pole than to the left pole).

Now let $m$ be the midpoint of $\ell$ and $r$. Let $d$ denote the distance from $q$ to the perpendicular bisector of $\overline{lr}$, as shown in Figure [ADD FIGURE]. Since we assign a point in the parent cluster to the left child if it is closer to the left pole, if $d > \rho$, (i.e., if $d$ is to the left of the perpendicular bisector of $\overline{\ell r}$), 
the left child does not overlap the query ball. In this case, we proceed only with the right child. If $d \leq \rho$, 
then both child clusters overlap the query ball.

We can do a quick check using values computed at partition time to determine whether $d \leq \rho$.

\begin{algorithm} 
    \caption{$\rho$-NN(\emph{cluster, query, r})} 
    \label{alg:methods:rnn-search} 
    \begin{algorithmic}
        \REQUIRE $r \geq 0$
        \REQUIRE $cluster \neq \emptyset$
        \STATE $results \Leftarrow \emptyset$
        \IF{$cluster.left \cap B_X(query, r) \neq \emptyset$}
            \IF{$cluster.left.\delta$ $\leq$ $r$ + $cluster.left.radius$}
                \STATE{$\rho$-NN($cluster.left, query, r$)}
            \ENDIF
        \ENDIF
        \IF{$cluster.right \cap B_X(query, r) \neq \emptyset$}
            \IF{$cluster.right.\delta$ $\leq$ $r$ + $cluster.right.radius$} 
                \STATE $\rho$-NN($cluster.right, query, r$)
            \ENDIF
        \ENDIF
        \IF{$cluster.isLeaf$}
            \FOR{$p \in cluster.points$}
                \IF{$p.\delta \leq r$}
                    \STATE $results.push((p, r))$
                \ENDIF
            \ENDFOR
        \ENDIF
        \STATE Return $results$
    \end{algorithmic}
\end{algorithm}

@@

First, we aim to find the element $q_{\tau} \in Q$ with the smallest $\delta_{\max}$ such that 
$q_{\tau}$ and all elements in $Q$ with smaller $\delta_{\max}$ collectively contain at least $k$ points.

To find this element, we use a process similar to the Partition algorithm used in Quicksort~\cite{10.1093/comjnl/5.1.10}, adjusted to account for the varying cardinalities of clusters in $Q$ by having each cluster count a number of times equal to its cardinality.
Our modified version of this algorithm finds the smallest index $i$ such that all elements in $Q$ with $\delta_{\max}$ less than or equal to $Q[i]$'s $\delta_{\max}$ collectively have cardinality greater than or equal to $k$.
We consider $Q[i]$ to be our $q_{\tau}$, and our threshold $\tau$ to be $Q[i]$'s $\delta_{\max}$.

We then remove from $Q$ any element whose $\delta_{\min}$ is greater than $\tau$.
Then we remove all leaf clusters from $Q$ and add their points to $Q$ instead.
Finally, we replace all non-leaf clusters in $Q$ with their children.

@@

\begin{algorithm} % enter the algorithm environment
    \caption{Sieve(\emph{root, query, k})} % give the algorithm a caption
    \label{alg:methods:sieve} % and a label for \ref{} commands later in the document
    \begin{algorithmic} % enter the algorithmic environment
        \REQUIRE $root$, $query$, $k$
        \REQUIRE $root.cardinality \geq k$
        \STATE $Q \Leftarrow$ [$root$]
        \WHILE{$|Q| > k$}
            \STATE $i \Leftarrow QuicksortPartition(Q, k)$
            \STATE $\tau \Leftarrow Q[i].\delta_{max}$
            \FOR {$q \in Q$}
                \IF {$q.\delta_{min} > \tau$}
                    \STATE $Q.pop(q)$
                \ENDIF
            \ENDFOR
            \FOR {$q \in Q$}
                \IF {$q.isLeaf \lor |q| \leq k$}
                    \STATE $Q.append(q.points)$
                \ELSE
                    \STATE $[l, r] \Leftarrow q.children$
                    \STATE $Q.append([l, r])$   
                \ENDIF
                \STATE $Q.remove(q)$
            \ENDFOR 
        \ENDWHILE
        \STATE Return $Q$
    \end{algorithmic}How do we algorithmically determine the depth to which we
\end{algorithm}

@@

We first let $Q$ be a min-queue of clusters prioritized by $\delta_{\min}$.
$Q$ starts containing only the root cluster.
We also let $H$ be a fixed-size ($k$) max-queue of points prioritized by $\delta$.
$H$ starts empty.

While $H$ has fewer than $k$ points or $\delta$ of the top priority element in $H$ is greater than $\delta_{\min}$ of the top priority element in $Q$, we do the following:

@@

\begin{algorithm} 
\caption{Greedy Sieve(\emph{root, query, k})} 
\label{alg:greedy-sieve} 
\begin{algorithmic}
    \STATE{$Q \Leftarrow$ priority queue}
    \STATE{$Q.push(root)$}
    \STATE{$H \Leftarrow$ priority queue}
    \WHILE{$|H| < k \lor H.peek.\delta > Q.peek.\delta_{\min}$}
        \WHILE{$\neg Q.peek.isLeaf$}
            \STATE{$[l, r] \Leftarrow Q.pop().children$}
            \STATE{$Q.push([l, r])$}
        \ENDWHILE
        \STATE{$leaf \Leftarrow Q.pop()$}
        \STATE{$H.push(leaf)$}
        \WHILE{$|H| > k$}
            \STATE{$H.pop()$}
        \ENDWHILE
    \ENDWHILE
    \STATE Return $H$
\end{algorithmic}
\end{algorithm}
